---
title: "🦋 The Butterfly Effect of Morning Delays (nycflights13)"
description: "Do small morning delays snowball into larger afternoon delays on the same day?"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    number-sections: false
---

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(nycflights13)
library(broom)
theme_set(theme_minimal(base_size = 13))
```

## Idea
Airports are complex systems. If departures are delayed **early** (5–9 a.m.), do those delays **propagate** into the **afternoon** (12–6 p.m.) the *same day*?

We’ll build a **daily panel** and look for same-day relationships, then examine **day-of-week** patterns and **within-day accumulation**.

## Build the daily panel

```{r}
period_of <- function(dep_time) {
  h <- floor(dep_time / 100)
  case_when(
    !is.na(h) & h >= 5  & h < 9  ~ "am_5_9",
    !is.na(h) & h >= 12 & h < 18 ~ "pm_12_18",
    TRUE ~ NA_character_
  )
}

daily <- flights |>
  mutate(period = period_of(dep_time)) |>
  filter(!is.na(period)) |>
  group_by(year, month, day, period) |>
  summarise(avg_dep_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = period, values_from = avg_dep_delay) |>
  mutate(
    date = as.Date(sprintf("%04d-%02d-%02d", year, month, day)),
    wday = lubridate::wday(date, label = TRUE, week_start = 1)
  ) |>
  rename(delay_am = am_5_9, delay_pm = pm_12_18) |>
  filter(is.finite(delay_am), is.finite(delay_pm))
```

## Do morning delays predict afternoon delays?

```{r}
p_scatter <- daily |>
  ggplot(aes(delay_am, delay_pm)) +
  geom_point(alpha = 0.55) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Avg departure delay (5–9 a.m., minutes)",
    y = "Avg departure delay (12–6 p.m., minutes)",
    title = "Morning delays are associated with worse afternoons (same day)"
  )
p_scatter
```

### Effect size (slope, R²)

```{r}
fit <- lm(delay_pm ~ delay_am, data = daily)
broom::tidy(fit)
broom::glance(fit) |> dplyr::select(r.squared, adj.r.squared)
```

**Interpretation.** A positive slope indicates **same-day propagation**: when the airport starts behind, it tends to **stay behind**.

## Does it vary by weekday?

```{r}
p_wday <- daily |>
  ggplot(aes(delay_am, delay_pm, color = wday)) +
  geom_point(alpha = 0.45) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  labs(
    title = "Propagation differs by weekday",
    x = "Morning delay (5–9 a.m.)", y = "Afternoon delay (12–6 p.m.)",
    color = "Weekday"
  )
p_wday
```

**Reading the plot.** Steeper lines imply **stronger carry-through** from morning to afternoon. If weekends look flatter, demand/traffic complexity may be lower.

## Within-day accumulation pattern

We check **hourly** averages to see if the day “drifts” upward after a bad start.

```{r}
hourly <- flights |>
  mutate(hour = floor(dep_time/100)) |>
  filter(!is.na(hour), hour >= 5, hour < 18) |>
  group_by(year, month, day, hour) |>
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE), .groups="drop") |>
  group_by(hour) |>
  summarise(p50 = median(avg_delay, na.rm = TRUE),
            p80 = quantile(avg_delay, 0.8, na.rm = TRUE),
            p20 = quantile(avg_delay, 0.2, na.rm = TRUE))

hourly |>
  ggplot(aes(hour, p50)) +
  geom_ribbon(aes(ymin = p20, ymax = p80), alpha = 0.15) +
  geom_line(linewidth = 1) +
  scale_x_continuous(breaks = seq(5,17,2)) +
  labs(
    title = "Median hourly delay with 20–80% band (5 a.m.–5 p.m.)",
    x = "Hour of day (departure)", y = "Delay (minutes)"
  )
```

**Takeaway.** If the band lifts after ~9–10 a.m., that suggests **accumulation**: early disruptions ripple forward as the schedule tightens.

## Summary
- **Yes:** Morning delays are **predictive** of afternoon delays the same day.  
- The effect **varies by weekday** (operational mix matters).  
- Hourly medians hint at **drift upward** after a rocky morning.

**Questions to explore next:** Are certain **carriers** or **routes** better at dampening propagation? Do **turnaround times** or **gate constraints** explain differences?

---

<footer>
**Story by Insightful Tales**  
_Last rendered on: `r format(Sys.Date(), '%B %d, %Y')`_
</footer>
