---
title: "The Geography of Luck"
subtitle: "The fortune of birthplaces and borders — tracing how chance and geography conspire to shape our destinies."
description: "What we call “luck” often begins with latitude and longitude. This story uncovers how place, mobility, and systems intertwine to shape the fortunes of people and nations."
keywords: [
  geography of luck,
  birthplace inequality,
  geographic inequality,
  borders and opportunity,
  mobility and place,
  global inequality,
  chance of birth,
  latitude effects,
  place and destiny,
  economic geography,
  spatial inequality,
  human development patterns,
  global mobility
]
categories: [geography, sociology, economics]
format:
  html:
    code-fold: show
    smooth-scroll: true
    df-print: paged
    anchor-sections: true
execute:
  warning: false
  message: false
  echo: true
  freeze: auto
---

<section class="hero">

<h1>🧭 The Geography of Luck</h1>

<h2 class="hero-subtitle">The fortune of birthplaces and borders — tracing how chance and geography conspire to shape our destinies.</h2>

<p class="hero-text">What we call “luck” often begins with latitude and longitude. This story uncovers how place, mobility, and systems intertwine to shape the fortunes of people and nations.</p>

</section>

> “Where the sun sits higher, shadows of inequality grow shorter — but never quite disappear.”

## [💡]{aria-hidden="true"} Premise

If you were to spin a globe and stop it at random, the odds of landing somewhere prosperous are not uniform. Latitude has long acted as the silent architect of opportunity — shaping climate, agriculture, disease, and, through them, destiny.\
This story asks: *Does geography still rule luck in the 21st century?*

------------------------------------------------------------------------

## [🌐]{aria-hidden="true"} Data Sources

We combine multiple open datasets (2020 snapshot for a clean cross-section):

| Theme | Source | Indicator |
|-----------------------:|:-----------------------|:-----------------------|
| Economic output | World Bank | GDP per capita (constant USD, `NY.GDP.PCAP.KD`) |
| Health | World Bank | Life expectancy at birth (`SP.DYN.LE00.IN`) |
| Geography | Natural Earth | Country boundaries + ISO3 (for joins) |

```{r}
#| label: setup
#| include: false
library(WDI); library(tidyverse); library(janitor); library(scales); library(readr)
# Spatial libs lazily, handle absence gracefully
has_ne <- requireNamespace("rnaturalearth", quietly = TRUE) &&
          requireNamespace("rnaturalearthdata", quietly = TRUE)
if (has_ne) { library(rnaturalearth); library(sf) }

# Theme
theme_set(theme_minimal(base_size = 13))

# --- World Bank snapshot (2020) ---
wb <- WDI(
  indicator = c(gdp = "NY.GDP.PCAP.KD", life = "SP.DYN.LE00.IN"),
  start = 2020, end = 2020, extra = TRUE
) |> clean_names()

# Ensure latitude numeric before using abs()
wb <- wb |> 
  dplyr::filter(region != "Aggregates") |>
  mutate(latitude = suppressWarnings(parse_number(as.character(latitude))),
         abs_lat  = abs(latitude),
         gdp_log  = log10(gdp))

# --- World polygons (optional) ---
if (has_ne) {
  world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") |>
    sf::st_as_sf() |>
    dplyr::select(iso_a3, name_long, continent, geometry)
  dat <- dplyr::left_join(world, wb, by = c("iso_a3" = "iso3c"))
} else {
  world <- NULL; dat <- NULL
  message("Optional map disabled: install.packages(c('rnaturalearth','rnaturalearthdata','sf')) to enable the choropleth.")
}
```

------------------------------------------------------------------------

## [🌱]{aria-hidden="true"} First Glance — The Shape of Prosperity

```{r}
# GDP vs |latitude|
ggplot(wb, aes(abs_lat, gdp)) +
  geom_point(aes(color = region), alpha = .7, size = 2, show.legend = FALSE) +
  geom_smooth(method = "loess", se = FALSE, color = "black") +
  scale_y_log10(labels = label_dollar()) +
  labs(title = "The Geography of Luck",
       subtitle = "GDP per capita (log scale) vs. absolute latitude, 2020 (World Bank)",
       x = "Absolute Latitude (°)", y = "GDP per capita (2015 USD, log scale)")
```

*Reading:* Prosperity tends to rise away from the equator, plateau in the mid‑latitudes, and soften again near the poles — a climatic parabola of fortune.

------------------------------------------------------------------------

## [💰]{aria-hidden="true"} Beyond Wealth — Life Follows Latitude

```{r}
# Life expectancy vs |latitude|
wb |> ggplot(aes(abs_lat, life)) +
  geom_point(alpha = .6, size = 2, color = "#555") +
  geom_smooth(method = "loess", se = FALSE, color = "#111") +
  labs(title = "Latitude vs. Life Expectancy",
       subtitle = "Life expectancy at birth vs. absolute latitude, 2020",
       x = "Absolute Latitude (°)", y = "Years")
```

*Reading:* Health shadows wealth: life expectancy arcs with latitude, but with notable outliers.

------------------------------------------------------------------------

## [🌍]{aria-hidden="true"} Exceptions — When Geography Loses (Residuals)

We fit a simple model of wealth on latitude and look for countries that **outperform** what geography alone would predict.

```{r}
# Fit on complete cases only, then join residuals back by iso3c
wb_cc <- wb |> dplyr::filter(is.finite(gdp_log), is.finite(abs_lat), !is.na(iso3c))
model_simple <- lm(gdp_log ~ abs_lat, data = wb_cc)
res_tbl <- tibble::tibble(iso3c = wb_cc$iso3c, resid_gdp_lat = unname(resid(model_simple)))

wb <- wb |> dplyr::left_join(res_tbl, by = "iso3c") |>
  dplyr::mutate(luck_index = as.numeric(scale(resid_gdp_lat)))

# Top/bottom outperformers by residuals (ignore NA)
out_top <- wb |> dplyr::filter(!is.na(resid_gdp_lat)) |>
  dplyr::slice_max(resid_gdp_lat, n = 10) |>
  dplyr::select(country, region, resid_gdp_lat)

out_bot <- wb |> dplyr::filter(!is.na(resid_gdp_lat)) |>
  dplyr::slice_min(resid_gdp_lat, n = 10) |>
  dplyr::select(country, region, resid_gdp_lat)

knitr::kable(out_top, digits = 2, caption = "Top 10 'Lucky Defiers' — richer than latitude alone predicts (log residual)")
```

```{r}
knitr::kable(out_bot, digits = 2, caption = "Bottom 10 — poorer than latitude alone predicts (log residual)")
```

> *Interpretation:* High‑performers often pair trade centrality, education, or resource rents with policy stability. Under‑performers frequently face governance or conflict frictions.

------------------------------------------------------------------------

## [🗺️]{aria-hidden="true"} Optional Map — The Luck Index (Residual Choropleth)

A **choropleth map** of the residuals ("luck index") highlights countries that over‑ or under‑perform relative to their absolute latitude.

```{r}
# Luck-index choropleth (guarded)
if (isTRUE(has_ne)) {
  # Merge residuals back to geometry
  map_dat <- dat |> dplyr::left_join(wb |> dplyr::select(iso3c, resid_gdp_lat), by = c("iso_a3" = "iso3c"))

  # Breaks for a symmetric diverging map around zero residual
  brks <- c(-Inf, -0.6, -0.3, -0.15, 0, 0.15, 0.3, 0.6, Inf)
  map_dat$res_bin <- cut(map_dat$resid_gdp_lat, breaks = brks, include.lowest = TRUE)

  # Plot
  ggplot(map_dat) +
    geom_sf(aes(fill = res_bin), color = "white", size = 0.1) +
    scale_fill_brewer(type = "div", palette = "RdBu", direction = -1, na.value = "#e5e7eb",
                      name = "Luck index
(GDP~vs~|lat| residual)") +
    labs(title = "Outperformers and Underperformers — Geography of Luck",
         subtitle = "Residuals from log(GDP per capita) ~ absolute latitude, 2020",
         caption = "Data: World Bank (WDI), Natural Earth. Negative = underperforming latitude; positive = outperforming.") +
    theme(legend.position = "right")
} else {
  message("Optional map disabled: install.packages(c('sf','rnaturalearth','rnaturalearthdata')) to enable the choropleth.")
}
```

*How to read:* **Blue** shades underperform their latitude (negative residuals); **red** shades outperform (positive residuals). Neutral grays are missing data.

------------------------------------------------------------------------

## [💭]{aria-hidden="true"} Reflection

> *The sun may rise for everyone, but it still shines longer for some.*

Latitude once dictated agriculture, disease, and labor. Today, it still whispers through economies — a relic of environmental inheritance.\
Yet each bright outlier on the map is a defiance of fate: policy and ingenuity turning geography into geography’s undoing.

------------------------------------------------------------------------

## [⚙️]{aria-hidden="true"} Technical Appendix

-   **Packages:** `WDI`, `rnaturalearth`, `sf`, `tidyverse`, `janitor`, `scales`\
-   **Year:** 2020 snapshot (World Bank)\
-   **Transforms:** `log10(GDP per capita)`; residuals from `lm(gdp_log ~ abs_lat)`\
-   **CRS:** EPSG:4326 (WGS84)\
-   **Reproducibility:** Set a specific year for comparability; extend to multi‑year to track weakening/strengthening latitude effects.

```{r}
# Time evolution — correlation weakening/strengthening over time (1980–2023)
hist <- WDI(indicator = c(gdp = "NY.GDP.PCAP.KD"),
            start = 1980, end = 2023, extra = TRUE) |>
  janitor::clean_names() |>
  dplyr::filter(region != "Aggregates") |>
  dplyr::mutate(
    # make sure latitude is numeric before abs()
    latitude = suppressWarnings(readr::parse_number(as.character(latitude))),
    abs_lat  = abs(latitude),
    gdp_log  = log10(gdp)
  ) |>
  dplyr::group_by(year) |>
  dplyr::summarize(
    r_lat_gdp = cor(abs_lat, gdp_log, use = "complete.obs"),
    .groups = "drop"
  )

ggplot(hist, aes(year, r_lat_gdp)) +
  geom_line() +
  labs(
    title = "Latitude–Wealth Correlation Over Time",
    x = "Year", y = "Correlation r (|lat| vs log GDP per capita)"
  )
```
