---
title: "🌆 Lights and Lives — Cities from Space"
subtitle: "Night-light data reveals how urban sprawl, inequality, and energy intersect."
description: "At night, Earth turns into a map of human presence. Street grids, ports, industrial belts,and unlit gaps sketch a world where energy, wealth, and opportunity are unevenly distributed. This story reads night-light data as a living atlas of cities — where luminosity becomes a proxy for power, access, and the shape of urban sprawl."
categories: ["urban-life", "climate-and-weather", "data-science", "global-trends"]
keywords: ["night lights", "VIIRS", "DMSP-OLS", "urban sprawl", "inequality", "energy use", "luminosity", "remote sensing", "city lights", "global cities"]
execute:
  eval: false
---

<section class="hero">

<h1>🌆 Lights and Lives — Cities from Space</h1>

<h2 class="hero-subtitle">

Night-light data reveals how urban sprawl, inequality, and energy intersect.

</h2>

<p class="hero-text">

At night, Earth turns into a map of human presence. Street grids, ports, industrial belts,and unlit gaps sketch a world where energy, wealth, and opportunity are unevenly distributed. This story reads night-light data as a living atlas of cities — where luminosity becomes a proxy for power, access, and the shape of urban sprawl.

</p>

</section>

::: {.callout-note appearance="minimal"}
**How to read this page:**\
We treat night-time lights as a proxy for human activity — not a perfect measure, but a powerful lens. Each chart is fully reproducible: adjust filters, swap cities, or plug in your own preprocessed night-light data.
:::

## 🌍 1. A planet that glows unevenly

From orbit, the Earth at night is both beautiful and unsettling.

Coastlines trace themselves in neon. Highways appear as filament veins. Some regions blaze so brightly that their urban boundaries blur; others — home to millions — barely register at all.

Night-time satellite imagery turns one simple question into a complex one:

> **Where do we put light, and what does that say about how we share energy, wealth, and space?**

In this story, we use processed night-light data to look at cities as luminous organisms:

-   How bright are different cities relative to their population?
-   How unevenly is light distributed within a metropolitan area?
-   What does the **shape** of light — tight and compact vs diffuse and sprawling — reveal about urban form?
-   Where are cities brightening, dimming, or fracturing into new luminous islands?

------------------------------------------------------------------------

## 💡 2. What night-light data actually measures

### 2.1 From radiance to a human-readable signal

Night-light datasets typically come from sensors like:

-   **DMSP-OLS** (historic, coarser data, 1990s–2013)
-   **VIIRS Day/Night Band (DNB)** (finer resolution, ongoing)

These sensors measure **radiance** — roughly, how much visible/near-visible light is emitted or reflected from each patch of the Earth.

But raw radiance is messy:

-   Gas flares are extremely bright and not “city life.”
-   Snow, clouds, and moonlight can contaminate readings.
-   Not all lights mean prosperity: prisons, industrial zones, and isolated highways glare without necessarily improving lives.

So we usually work with **processed composites**:

-   Cloud-masked
-   Averaged over months or years
-   With ephemeral or non-urban light sources suppressed where possible

### 2.2 The dataset we’ll use in this story

For this story, we imagine you’ve prepared two tidy datasets:

1.  **City-year summary**

```{r}
library(tidyverse)
library(lubridate)

night_cities <- readr::read_csv("data/nightlights_city_year.csv")
glimpse(night_cities)
```

```         
Expected columns (one row per city-year):

- `city` — City name (e.g., "Lagos", "Chicago")
- `country` — Country name
- `region` — Broad region (e.g., "Africa", "Europe", "Asia")
- `year` — Year of observation (numeric)
- `population_millions` — Estimated metro population (millions)
- `gdp_per_capita` — GDP per capita (constant USD, optional)
- `mean_radiance` — Mean night-light radiance within a metro footprint
- `total_radiance` — Sum of radiance across the metro area
- `lit_area_km2` — Area above a brightness threshold (sq km)
- `core_share` — Share of radiance in the brightest 20% of pixels (compactness vs sprawl)
- `coast_distance_km` — Distance of city center to nearest coastline (optional)
```

2.  **Radial profiles for sprawl**

```{r}
night_radial <- readr::read_csv("data/nightlights_radial_profiles.csv")
glimpse(night_radial)
```

Expected columns (multiple rings per city-year):

-   `city`, `country`, `year`
-   `distance_km` — Distance from city center (e.g., 0–5 km, 5–10 km, …)
-   `mean_radiance_ring` — Mean radiance in that ring
-   `population_ring` — Estimated population in that ring (optional)

These structures let us ask three big questions:

1.  **How bright are cities, given their size?**
2.  **How uneven is the distribution of light within a city?**
3.  **How far does the glow of a city extend into the dark?**

------------------------------------------------------------------------

## 🏙️ 3. Brightness vs size: are big cities always brighter?

If night-time lights correlate with economic activity, we might expect a simple relationship:

> Bigger city → more people → more light.

In practice, the relationship is noisy. Let’s look at how mean brightness scales with population across regions.

```{r}
library(ggplot2)

# Focus on a recent snapshot
recent_year <- max(night_cities$year, na.rm = TRUE)

city_recent <- night_cities |>
  filter(year == recent_year,
         population_millions > 0.5) |>  # avoid tiny settlements
  mutate(region = forcats::fct_reorder(region, mean_radiance, .fun = median))

ggplot(city_recent,
       aes(x = population_millions, y = mean_radiance)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ region) +
  labs(
    title = glue::glue("How Bright Are Big Cities? ({recent_year})"),
    subtitle = "Log–log relationship between population and mean night-time radiance, by region.",
    x = "Metro population (millions, log scale)",
    y = "Mean night-time radiance (log scale)",
    caption = "Each point is a metro area; lines show simple linear fits on log–log scales."
  ) +
  theme_minimal(base_size = 13)
```

We typically see:

-   **Europe & East Asia**: Large cities with high, relatively consistent brightness.
-   **South Asia & Africa**: Rapidly growing cities where population outpaces luminous infrastructure.
-   **North America**: Bright, sprawling metros with high radiance even at modest population sizes.

In some regions, being a **big city** doesn’t guarantee a bright, evenly lit urban fabric. Light becomes a measure of **how much infrastructure accompanies growth**, not just growth itself.

## ⚖️ 4. Unequal light, unequal lives

One of the most revealing questions we can ask is:

> Within a city, how much of the light belongs to how few of the people?

We can approximate this by comparing cumulative population and cumulative radiance across neighborhoods or pixels.

Assume we’ve precomputed a **“light Lorenz”** summary per city:

```{r}
light_lorenz <- readr::read_csv("data/nightlights_lorenz_city.csv")
glimpse(light_lorenz)
```

Expected columns:

-   `city`, `country`, `year`
-   `cum_pop_share` — Cumulative share of population (0 → 1)
-   `cum_light_share` — Cumulative share of radiance (0 → 1)
-   `region`

Let’s compare a few cities with distinct urban stories:

```{r}
cities_focus <- c("Lagos", "Paris", "Los Angeles", "Mumbai")

lorenz_focus <- light_lorenz |>
  filter(year == recent_year,
         city %in% cities_focus)

ggplot(lorenz_focus,
       aes(x = cum_pop_share, y = cum_light_share, color = city)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_line(linewidth = 1.1) +
  coord_equal() +
  labs(
    title = glue::glue("Unequal Light Within Cities ({recent_year})"),
    subtitle = "Cumulative share of night-time light vs cumulative share of population.",
    x = "Cumulative share of population",
    y = "Cumulative share of night-time light",
    caption = "The farther a curve bows below the diagonal, the more concentrated the light is in a small share of space/people."
  ) +
  theme_minimal(base_size = 13)
```

Reading this chart:

-   A city whose curve stays **close to the diagonal** has a more even distribution of light.
-   A city with a curve that **bows sharply** indicates that a small fraction of space holds most of the light — dense, concentrated cores and dark peripheries.
-   Comparing multiple cities becomes a way of comparing **infrastructural inequality**, not just income inequality.

Night-time lights can’t tell us everything about welfare. But they can reveal where the urban fabric is **visibly uneven**.

## 🌐 5. The shape of sprawl: how far does a city’s glow reach?

Some cities are tight clusters of light with a sharp edge. Others dissolve gradually into faint, far-reaching halos of suburbia and exurban development.

We can visualize this using **radial profiles** — how brightness decays with distance from the city center.

```{r}
cities_sprawl <- c("Los Angeles", "Paris", "Shanghai", "Johannesburg")

radial_recent <- night_radial |>
  filter(year == recent_year,
         city %in% cities_sprawl)

ggplot(radial_recent,
       aes(x = distance_km, y = mean_radiance_ring, color = city)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = glue::glue("The Glow Radius of Cities ({recent_year})"),
    subtitle = "Average night-time radiance as a function of distance from city center.",
    x = "Distance from city center (km)",
    y = "Mean radiance in radial ring",
    caption = "Long, shallow tails suggest extensive sprawl; sharp drop-offs indicate more compact light footprints."
  ) +
  theme_minimal(base_size = 13)
```

Interpretation:

-   A **slow decay** curve indicates a long, fuzzy urban edge — the light keeps going.
-   A **steep drop-off** suggests a clear urban boundary, perhaps due to planning, geography, or policy.
-   Two cities with similar populations can have radically different **spatial energy signatures**.

If we overlay population by ring, we can even ask: *Where is the mismatch between where people live and where the light is?*

## 📈 6. Cities that brighten, cities that dim

Lights don’t just show space — they show **change over time**.

Some cities brighten steadily as the grid extends and incomes rise. Others brighten at the edges while their cores dim, hinting at shifting activity patterns, suburbanization, or deindustrialization.

Let’s look at trajectories for a diverse set of cities:

```{r}
cities_time <- c("Lagos", "Dhaka", "Chicago", "Berlin", "São Paulo")

city_trends <- night_cities |>
  filter(city %in% cities_time)

ggplot(city_trends,
       aes(x = year, y = mean_radiance, color = city)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.3) +
  labs(
    title = "Brightening and Dimming Cities Over Time",
    subtitle = "Mean night-time radiance for selected metros, across years.",
    x = "Year",
    y = "Mean radiance",
    caption = "Upward trends often reflect electrification and economic growth; plateaus or declines may signal saturation, efficiency gains, or changing land use."
  ) +
  theme_minimal(base_size = 13)
```

We might see:

-   **Explosive brightening** in emerging megacities.
-   **Smoother, slower growth** in already wealthy metros.
-   **Flat or declining trends** where industries have moved, or where cities are densifying without increasing light.

You can extend this to GDP per capita, electrification rates, or emissions — using light as a visual companion to conventional indicators.

## 🔌 7. Light per person: a rough currency of energy access

If we divide total radiance by population, we get a rough metric:

> **Light per person** = total_radiance / population

It’s far from perfect, but as a comparative index it can be revealing.

```{r}
intensity_recent <- city_recent |>
  mutate(light_per_person = total_radiance / (population_millions * 1e6))

ggplot(intensity_recent,
       aes(x = gdp_per_capita, y = light_per_person, color = region)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.7) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = glue::glue("Light per Person vs Income ({recent_year})"),
    subtitle = "A rough proxy for the intersection of energy, infrastructure, and economic status.",
    x = "GDP per capita (log scale)",
    y = "Estimated light per person (log scale)",
    caption = "Higher values suggest more luminous infrastructure per resident; interpretation requires care around efficiency and patterns of land use."
  ) +
  theme_minimal(base_size = 13)
```

Patterns to watch:

-   Regions where income rises faster than light: efficiency gains, dense housing, stricter lighting policies.
-   Regions where light explodes with modest income gains: rapid grid expansion, car-oriented sprawl, or energy-intensive industries.

Light here becomes a crude but evocative **currency of visibility and access**.

## 🚫 8. What lights can’t see (and why that matters)

The seduction of night-time imagery is its simplicity: bright means “developed”, dark means “left behind.” Reality is messier.

Night lights **cannot** directly see:

-   The quality of housing or labor conditions.
-   Whether energy is clean or fossil-derived.
-   Whether bright light is freedom (safe streets) or surveillance (floodlit walls).
-   Quiet forms of wealth — digital economies, services, and high-value work in dim offices.

What they *do* capture extraordinarily well is **the footprint of energy-intensive infrastructure** — roads, ports, factories, malls, and dense residential grids.

So the right question becomes:

> Not “Is this place developed?” but “What **kind** of development does this luminous footprint suggest?”

## ✨ 9. Lights as a vocabulary of urban form

Satellite night lights give us a new vocabulary for cities:

-   **Sharp-edged cores** vs **soft halos**
-   **Single dominant cluster** vs **polycentric constellations**
-   **Linear corridors** of light along rivers, coasts, or highways
-   **Isolated flares** of industrial complexes and ports

You can imagine a future where cities are described not just by population and GDP, but by their:

-   **Luminous compactness**
-   **Sprawl tail length**
-   **Light–population inequality index**
-   **Temporal brightening profile**

Each of these becomes a way to compare cities as living systems.

## ➡️ 10. Where to go from here

This story barely scratches the surface of what’s possible with night-light data.

Ideas for further exploration:

-   **Resilience:** How do lights flicker and recover around disasters, blackouts, or conflicts?
-   **Seasonality:** How do festive periods or monsoon seasons change the luminous texture of regions?
-   **Policy impacts:** Before/after analyses of electrification programs, zoning reforms, or efficiency standards.
-   **Climate:** Linking luminous sprawl to commuting patterns, vehicle miles traveled, or emissions.

Night-time lights are not a perfect proxy for human well-being. But they are one of the few global, consistent, high-frequency measures we have of **how we inhabit the dark**.

<section class="dataset-source">
  <img src="../images/dataset-icons/night-lights.svg"
       class="dataset-icon-small"
       alt=""
       aria-hidden="true">
  <span><strong>Dataset:</strong> 
    <a href="../datasets/nasa-noaa-night-time-lights-earths-nocturnal-footprint-seen-from-space.html">NASA / NOAA Night-Time Lights</a>
  </span>
</section>

<p class="dataset-subnav">
  <a href="../dataset-icon-language.html" class="small-link">
    What do these icons mean?
  </a>
</p>

## 📚 References and credits

-   **NASA / NOAA VIIRS Night-time Lights** — Day/Night Band composites processed for cloud masking and stray-light filtering.
-   **DMSP-OLS Night-time Lights** — Historic low-light imaging of Earth at night, 1990s–2013.
-   Various research efforts on night lights and development, including studies on:
    -   Mapping electrification progress.
    -   Estimating economic activity in data-sparse regions.
    -   Quantifying urban sprawl and land-use change.

Data pre-processing steps — including cloud masking, monthly-to-annual compositing, thresholding of gas flares, and city footprint definitions — should be documented in your site’s dataset page for **NASA / NOAA Night Lights**.

## 🔗 See also on Insightful Tales

-   🌍 **Geography of Luck** — On the accidents of latitude, climate, and opportunity.
-   🌏 **When Growth Forgot Happiness** — A Gapminder journey through wealth and well-being.
-   🛰️ **Planetary Pulse** — Satellite time series of Earth’s rhythms in vegetation, snow, and fire.

Together, these stories form a quiet thesis:

> From orbit, the world’s inequalities are visible not just in charts, but in where the light falls — and where it does not.
