---
title: "🌦️ Weather Memory: Do Wind & Visibility in the Morning Predict Afternoon Delays?"
description: "Measuring 'memory' from morning weather (5–9 a.m.) to afternoon departure delays (12–6 p.m.)."
format:
  html:
    toc: true
    code-fold: true
---

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(nycflights13)
library(broom)
theme_set(theme_minimal(base_size = 13))
```

## Question
Does **morning weather** (5–9 a.m.) leave a measurable **imprint** on **afternoon (12–6 p.m.) departure delays**?

## Build a daily weather–delay panel

```{r}
# Flight delays by morning/afternoon
fl_daily <- flights |>
  filter(!is.na(dep_delay), !is.na(dep_time)) |>
  mutate(hour = floor(dep_time/100),
         period = case_when(
           hour >= 5  & hour < 9  ~ "am_5_9",
           hour >= 12 & hour < 18 ~ "pm_12_18",
           TRUE ~ NA_character_
         )) |>
  filter(!is.na(period)) |>
  group_by(year, month, day, period) |>
  summarise(avg_dep_delay = mean(dep_delay, na.rm = TRUE), .groups="drop") |>
  pivot_wider(names_from = period, values_from = avg_dep_delay)

# Morning weather aggregated across origins
wx_am <- weather |>
  filter(hour >= 5, hour < 9) |>
  group_by(year, month, day, origin) |>
  summarise(
    visib_am = mean(visib, na.rm=TRUE),
    wind_am  = mean(wind_speed, na.rm=TRUE),
    precip_am = mean(precip, na.rm=TRUE),
    .groups="drop"
  ) |>
  group_by(year, month, day) |>
  summarise(across(c(visib_am, wind_am, precip_am), ~mean(.x, na.rm=TRUE)), .groups="drop")

panel <- fl_daily |>
  inner_join(wx_am, by = c("year","month","day")) |>
  mutate(date = as.Date(sprintf("%04d-%02d-%02d", year, month, day))) |>
  rename(delay_am = am_5_9, delay_pm = pm_12_18) |>
  filter(is.finite(delay_pm), is.finite(delay_am))
```

## First look: simple associations

```{r}
panel |>
  pivot_longer(c(visib_am, wind_am, precip_am), names_to = "wx", values_to = "val") |>
  ggplot(aes(val, delay_pm)) +
  geom_point(alpha = 0.45) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ wx, scales = "free_x") +
  labs(
    title = "Morning weather vs. afternoon delays (same day)",
    x = "Morning weather metric", y = "Afternoon avg dep delay (min)"
  )
```

**Reading the facets.**  
- **wind_am ↑ → delay_pm ↑?** suggests wind-related capacity/resequencing effects.  
- **visib_am ↓ → delay_pm ↑?** supports low-visibility knock-on effects.  
- **precip_am** can capture storms/showers lingering into midday.

## Adjusted effects (multiple regression)

```{r}
fit <- lm(delay_pm ~ delay_am + wind_am + visib_am + precip_am, data = panel)
broom::tidy(fit)
broom::glance(fit) |> dplyr::select(r.squared, adj.r.squared, sigma)
```

**Interpretation.** Including **delay_am** controls for the morning’s operational state; remaining weather coefficients show **extra** afternoon impact beyond “we already started late.”

## Thresholds: when wind gets “bad”

```{r}
cut_wind <- median(panel$wind_am, na.rm = TRUE)
panel |>
  mutate(wind_bin = if_else(wind_am >= cut_wind, "Windy AM (≥ median)", "Calmer AM (< median)")) |>
  ggplot(aes(delay_am, delay_pm, color = wind_bin)) +
  geom_point(alpha = 0.45) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Propagation is worse on windier mornings",
    x = "Morning delay (5–9 a.m.)", y = "Afternoon delay (12–6 p.m.)", color = NULL
  )
```

**Takeaway.** On windier mornings, the **morning→afternoon slope** is typically steeper: the system absorbs **less** disruption.

## Summary
- Morning **wind**, **visibility**, and **precipitation** have **same-day** relationships with afternoon delays.  
- Even after controlling for **morning delay level**, weather still shows **additional** signal.  
- Windier mornings tend to **amplify** propagation (steeper slope).

**Next:** Add **origin-specific** effects, interact weather × carrier, or use **hour-level** models to separate schedule compression from weather persistence.

---

<footer>
**Story by Insightful Tales**  
_Last rendered on: `r format(Sys.Date(), '%B %d, %Y')`_
</footer>
